---
title: "HBCU vs PWI Environmental Justice Analysis: Full Reproducible Pipeline"
subtitle: "Data acquisition, processing, and statistical analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-summary: "Show code"
    theme: flatly
    fig-width: 10
    fig-height: 6
    embed-resources: true
    df-print: kable
execute:
  warning: false
  message: false
  cache: false
---

#  Notes

This document provides a **complete, self-contained pipeline** for reproducing all results in:

> Stanbrook, R. (in press). Environmental injustice in the shadow of the ivory tower: Cumulative EJ burden in communities surrounding Historically Black Colleges and Universities compared to primarily-white institutions.

The pipeline is structured in three parts:

1.  **Data Acquisition** — queries the EPA EJScreen ArcGIS REST API to retrieve EJ indicator values for campus communities, following the same 1-mile buffer / population-weighted aggregation approach used in the published analysis.
2.  **Data Processing** — tidies raw API output into the analysis-ready wide format CSVs archived in this repository.
3.  **Statistical Analysis** — reproduces all descriptive statistics, figures, and inferential tests reported in the paper.

## Using the archived data (recommended for exact replication)

The Dryad repository includes pre-computed, analysis-ready CSV files:

| File               | Contents                                        |
|--------------------|-------------------------------------------------|
| `hbcu_list.csv`    | HBCU campus centroids and metadata (n = 87)     |
| `pwi_list.csv`     | PWI campus centroids and metadata (n = 86)      |
| `hbcu_ej_wide.csv` | HBCU EJScreen EJ Index percentiles, wide format |
| `pwi_ej_wide.csv`  | PWI EJScreen EJ Index percentiles, wide format  |

To reproduce all **analysis results exactly**, set `USE_ARCHIVED_DATA <- TRUE` in the chunk below and ensure the four CSV files are in your working directory.

```{r global-config}
# ── SET THIS FLAG ──────────────────────────────────────────────────────────────
# TRUE  → skip API fetch; load archived CSVs from the repository deposit
# FALSE → re-fetch from the live EJScreen API (requires internet access)
USE_ARCHIVED_DATA <- TRUE

# EJScreen service version used in the published analysis
EJSCREEN_VERSION <- "EJScreen_2.2_with_AS_CNMI_GU_VI"

# Buffer radius in metres (1 US mile)
BUFFER_M <- 1609

# Set a seed for all bootstrapped CIs (must match published analysis)
RANDOM_SEED <- 42
```

------------------------------------------------------------------------

# Setup

## Required packages

```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)   # data wrangling and visualisation
  library(httr)        # HTTP requests to EJScreen API
  library(jsonlite)    # JSON parsing
  library(sf)          # spatial operations
  library(tigris)      # US state boundaries for maps
  library(leaflet)     # interactive maps
  library(MatchIt)     # propensity-score matching
  library(broom)       # tidy model output
  library(scales)      # axis formatting
  library(gt)          # publication-quality tables
  library(ggridges)    # ridgeline density plots
  library(ggrepel)     # non-overlapping map labels
  library(car)         # Levene's test
  library(readr)       # CSV I/O
  library(dplyr)
  library(stringr)
})

# car::recode masks dplyr::recode — restore dplyr's version
recode <- dplyr::recode

# Global ggplot theme
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey40"),
      panel.grid.minor = element_blank()
    )
)
```

To install all required packages from scratch:

``` r
install.packages(c(
  "tidyverse", "httr", "jsonlite", "sf", "tigris", "leaflet",
  "MatchIt", "broom", "scales", "gt", "ggridges", "ggrepel",
  "car", "readr", "dplyr", "stringr"
))
```

R version used in published analysis: **R 4.5.1**

------------------------------------------------------------------------

# Part 1: Data Acquisition

> Skip to [Part 2](#part-2-data-processing) if `USE_ARCHIVED_DATA = TRUE`.

## 1.1 Institution lists

### HBCU list

The HBCU institution list was compiled from a curated institutional dataset (`HBCU-DATA2.0.csv`) and cross-validated against federal Title III designation records. The raw file is included in the repository as `HBCU-DATA2.0.csv`. The code below replicates the cleaning applied by `scripts/prepare_hbcu_list.R`.

```{r prepare-hbcu-list, eval=!USE_ARCHIVED_DATA}
raw_hbcu <- readr::read_csv(
  "HBCU-DATA2.0.csv",
  col_names = c("school", "Location", "Latitude", "Longitude", "year_founded"),
  skip = 1,
  show_col_types = FALSE,
  trim_ws = TRUE
)

hbcu_list <- raw_hbcu |>
  # Drop empty rows
  dplyr::filter(!is.na(school), school != "") |>
  # Clean school names
  dplyr::mutate(
    school = stringr::str_remove_all(school, '"') |> stringr::str_squish()
  ) |>
  # Coerce coordinates; US longitudes must be negative
  dplyr::mutate(
    Latitude  = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    Longitude = dplyr::if_else(Longitude > 0, -Longitude, Longitude)
  ) |>
  # Remove records outside US/territory bounding box
  dplyr::filter(
    !is.na(Latitude),  Latitude  >= 15, Latitude  <= 72,
    !is.na(Longitude), Longitude <= -60, Longitude >= -180
  ) |>
  # Parse founding year
  dplyr::mutate(
    year_founded = as.integer(stringr::str_extract(
      as.character(year_founded), "\\d{4}"
    )),
    Location = stringr::str_squish(Location)
  ) |>
  dplyr::select(school, Latitude, Longitude, Location, year_founded) |>
  dplyr::arrange(school)

message(sprintf("HBCU list: %d institutions", nrow(hbcu_list)))
```

### PWI list

The PWI comparison group was assembled manually from publicly available federal data. Selection criteria: (i) four-year, degree-granting, accredited institutions classified as primarily-white under federal Title III definitions; (ii) located in a state also represented in the HBCU sample; (iii) comparable undergraduate enrollment; (iv) no specialist mission (military academies, seminaries excluded). The curated list is archived as `pwi_list.csv`.

```{r load-pwi-list, eval=!USE_ARCHIVED_DATA}
# pwi_list.csv is manually curated — load directly
pwi_list <- readr::read_csv("pwi_list.csv", show_col_types = FALSE)
message(sprintf("PWI list: %d institutions", nrow(pwi_list)))
```

## 1.2 EJScreen API helper functions

EJ burden data are retrieved from the EPA EJScreen ArcGIS Online FeatureServer. For each institution a spatial query is submitted using the campus centroid coordinates and a 1-mile (1,609 m) circular buffer. All census block groups whose boundaries intersect the buffer are returned.

```{r ejscreen-helpers, eval=!USE_ARCHIVED_DATA}
# ── EJScreen ArcGIS REST endpoint ─────────────────────────────────────────────
EJSCREEN_URL <- paste0(
  "https://services.arcgis.com/cJ9YHowT8TU7DUyn/ArcGIS/rest/services/",
  EJSCREEN_VERSION,
  "/FeatureServer/1/query"
)

# EJ Summary Index national (USA) percentile field names in EJScreen 2.2
# These are the raw API field names; they are renamed to descriptive labels
# in the processing step below.
EJ_FIELDS <- c(
  "P_EJDPM25SUP",  # Particulate Matter (PM2.5) EJ Index
  "P_EJOZONESUP",  # Ozone EJ Index
  "P_EJNO2SUP",    # Nitrogen Dioxide (NO2) EJ Index
  "P_EJDSLPMSUP",  # Diesel Particulate Matter EJ Index
  "P_EJCANCR",     # Toxic Releases to Air (Air Toxics Cancer Risk) EJ Index
  "P_EJPTRAF",     # Traffic Proximity and Volume EJ Index
  "P_EJLDPNT",     # Lead Paint EJ Index
  "P_EJPNPL",      # Superfund Proximity EJ Index
  "P_EJPRMP",      # RMP Facility Proximity EJ Index
  "P_EJPTSDF",     # Hazardous Waste Proximity EJ Index
  "P_EJUST",       # Underground Storage Tanks EJ Index
  "P_EJPWDIS",     # Wastewater Discharge EJ Index
  "P_EJDWATERNC",  # Drinking Water Non-Compliance EJ Index
  "ACSTOTPOP"      # Total resident population (for population weighting)
)

# Human-readable names used in all analyses — must match pu_vars in Part 3
EJ_LABELS <- c(
  "P_EJDPM25SUP"  = "Particulate Matter Summary Index_pct_usa",
  "P_EJOZONESUP"  = "Ozone Summary Index_pct_usa",
  "P_EJNO2SUP"    = "Nitrogen Dioxide (NO2) Summary Index_pct_usa",
  "P_EJDSLPMSUP"  = "Diesel Particulate Matter Summary Index_pct_usa",
  "P_EJCANCR"     = "Toxic Releases to Air Summary Index_pct_usa",
  "P_EJPTRAF"     = "Traffic Proximity and Volume Summary Index_pct_usa",
  "P_EJLDPNT"     = "Lead Paint Summary Index_pct_usa",
  "P_EJPNPL"      = "Superfund Proximity Summary Index_pct_usa",
  "P_EJPRMP"      = "RMP Proximity Summary Index_pct_usa",
  "P_EJPTSDF"     = "Hazardous Waste Proximity Summary Index_pct_usa",
  "P_EJUST"       = "Underground Storage Tanks Summary Index_pct_usa",
  "P_EJPWDIS"     = "Wastewater Discharge Summary Index_pct_usa",
  "P_EJDWATERNC"  = "Drinking Water Non-Compliance Summary Index_pct_usa"
)

# ── Core fetch function ────────────────────────────────────────────────────────
# Queries EJScreen for a single campus centroid, returns raw block-group rows.
fetch_ejscreen_campus <- function(school_name, lat, lon,
                                  buffer_m = BUFFER_M,
                                  max_retries = 3,
                                  pause_s = 1.5) {
  out_fields <- paste(EJ_FIELDS, collapse = ",")

  params <- list(
    geometry     = paste0(lon, ",", lat),
    geometryType = "esriGeometryPoint",
    inSR         = "4326",
    spatialRel   = "esriSpatialRelIntersects",
    distance     = buffer_m,
    units        = "esriSRUnit_Meter",
    outFields    = out_fields,
    returnGeometry = "false",
    f            = "json"
  )

  for (attempt in seq_len(max_retries)) {
    resp <- tryCatch(
      httr::GET(EJSCREEN_URL, query = params, httr::timeout(60)),
      error = function(e) NULL
    )

    if (!is.null(resp) && httr::status_code(resp) == 200) {
      body <- httr::content(resp, as = "text", encoding = "UTF-8")
      parsed <- jsonlite::fromJSON(body, simplifyDataFrame = TRUE)

      if (!is.null(parsed$features) && nrow(parsed$features) > 0) {
        attrs <- parsed$features$attributes
        attrs$school <- school_name
        Sys.sleep(pause_s)
        return(attrs)
      } else {
        # No intersecting block groups found — return empty row
        empty <- as.data.frame(
          setNames(
            as.list(rep(NA_real_, length(EJ_FIELDS))),
            EJ_FIELDS
          )
        )
        empty$school <- school_name
        return(empty)
      }
    }

    Sys.sleep(pause_s * attempt)
  }

  warning(sprintf("Failed to fetch data for: %s", school_name))
  return(NULL)
}
```

## 1.3 Fetch EJScreen data for all institutions

```{r fetch-hbcu, eval=!USE_ARCHIVED_DATA}
message("Fetching EJScreen data for HBCUs (this may take ~10 minutes)...")

hbcu_raw_list <- purrr::map2(
  hbcu_list$school,
  hbcu_list |> select(Latitude, Longitude) |> transpose(),
  function(school, coords) {
    fetch_ejscreen_campus(school, coords$Latitude, coords$Longitude)
  }
)

hbcu_ejscreen_raw <- dplyr::bind_rows(
  purrr::compact(hbcu_raw_list)
)

readr::write_csv(hbcu_ejscreen_raw, "hbcu_ejscreen_raw.csv")
message(sprintf("Fetched %d block-group rows for %d HBCU campuses",
                nrow(hbcu_ejscreen_raw),
                n_distinct(hbcu_ejscreen_raw$school)))
```

```{r fetch-pwi, eval=!USE_ARCHIVED_DATA}
message("Fetching EJScreen data for PWIs (this may take ~10 minutes)...")

pwi_raw_list <- purrr::map2(
  pwi_list$school,
  pwi_list |> select(Latitude, Longitude) |> transpose(),
  function(school, coords) {
    fetch_ejscreen_campus(school, coords$Latitude, coords$Longitude)
  }
)

pwi_ejscreen_raw <- dplyr::bind_rows(
  purrr::compact(pwi_raw_list)
)

readr::write_csv(pwi_ejscreen_raw, "pwi_ejscreen_raw.csv")
message(sprintf("Fetched %d block-group rows for %d PWI campuses",
                nrow(pwi_ejscreen_raw),
                n_distinct(pwi_ejscreen_raw$school)))
```

------------------------------------------------------------------------

# Part 2: Data Processing {#part-2-data-processing}

## 2.1 Population-weighted aggregation

For each campus, EJ index percentile values from all intersecting block groups are aggregated to a single campus-level estimate using a **population-weighted mean** (weighted by `ACSTOTPOP`, the resident population of each block group). Block groups with zero or missing population are excluded from the weighting but contribute a simple mean if no populated block groups exist.

```{r aggregate-function, eval=!USE_ARCHIVED_DATA}
# Computes the population-weighted mean of EJ indicators across block groups
# for a single campus.
aggregate_campus <- function(df_campus) {
  ej_cols <- names(EJ_LABELS)  # the 13 raw API field names

  pop <- df_campus$ACSTOTPOP
  pop[is.na(pop) | pop < 0] <- 0
  total_pop <- sum(pop)

  if (total_pop == 0) {
    # No populated block groups — fall back to simple mean
    result <- df_campus |>
      summarise(across(all_of(ej_cols), ~ mean(.x, na.rm = TRUE)))
  } else {
    result <- df_campus |>
      summarise(across(
        all_of(ej_cols),
        ~ weighted.mean(.x, w = pop, na.rm = TRUE)
      ))
  }

  result
}
```

```{r process-hbcu, eval=!USE_ARCHIVED_DATA}
hbcu_ejscreen_raw <- readr::read_csv("hbcu_ejscreen_raw.csv",
                                     show_col_types = FALSE)

hbcu_ej_wide <- hbcu_ejscreen_raw |>
  dplyr::group_by(school) |>
  dplyr::group_modify(~ aggregate_campus(.x)) |>
  dplyr::ungroup() |>
  # Rename API field names to descriptive labels used throughout analysis
  dplyr::rename(!!!EJ_LABELS) |>
  # Join metadata from institution list
  dplyr::left_join(
    hbcu_list |> select(school, Latitude, Longitude, Location, year_founded),
    by = "school"
  )

readr::write_csv(hbcu_ej_wide, "hbcu_ej_wide.csv")
message(sprintf("hbcu_ej_wide.csv written: %d rows", nrow(hbcu_ej_wide)))
```

```{r process-pwi, eval=!USE_ARCHIVED_DATA}
pwi_ejscreen_raw <- readr::read_csv("pwi_ejscreen_raw.csv",
                                    show_col_types = FALSE)

pwi_ej_wide <- pwi_ejscreen_raw |>
  dplyr::group_by(school) |>
  dplyr::group_modify(~ aggregate_campus(.x)) |>
  dplyr::ungroup() |>
  dplyr::rename(!!!EJ_LABELS) |>
  dplyr::left_join(
    pwi_list |> select(school, Latitude, Longitude, Location, year_founded),
    by = "school"
  )

readr::write_csv(pwi_ej_wide, "pwi_ej_wide.csv")
message(sprintf("pwi_ej_wide.csv written: %d rows", nrow(pwi_ej_wide)))
```

------------------------------------------------------------------------

# Part 3: Statistical Analysis

## 3.1 Load analysis-ready data

```{r load-data}
ej_hbcu <- readr::read_csv("hbcu_ej_wide.csv", show_col_types = FALSE)
ej_pwi  <- readr::read_csv("pwi_ej_wide.csv",  show_col_types = FALSE)
```

## 3.2 Variable definitions

```{r variable-defs}
# The 13 EJ Summary Index national (USA) percentile columns
pu_vars <- c(
  "Particulate Matter Summary Index_pct_usa",
  "Ozone Summary Index_pct_usa",
  "Nitrogen Dioxide (NO2) Summary Index_pct_usa",
  "Diesel Particulate Matter Summary Index_pct_usa",
  "Toxic Releases to Air Summary Index_pct_usa",
  "Traffic Proximity and Volume Summary Index_pct_usa",
  "Lead Paint Summary Index_pct_usa",
  "Superfund Proximity Summary Index_pct_usa",
  "RMP Proximity Summary Index_pct_usa",
  "Hazardous Waste Proximity Summary Index_pct_usa",
  "Underground Storage Tanks Summary Index_pct_usa",
  "Wastewater Discharge Summary Index_pct_usa",
  "Drinking Water Non-Compliance Summary Index_pct_usa"
)

# Short display labels for figures
indicator_labels <- c(
  "Particulate Matter Summary Index_pct_usa"          = "PM 2.5",
  "Ozone Summary Index_pct_usa"                       = "Ozone",
  "Nitrogen Dioxide (NO2) Summary Index_pct_usa"      = "NO₂",
  "Diesel Particulate Matter Summary Index_pct_usa"   = "Diesel PM",
  "Toxic Releases to Air Summary Index_pct_usa"       = "Toxic Air",
  "Traffic Proximity and Volume Summary Index_pct_usa"= "Traffic",
  "Lead Paint Summary Index_pct_usa"                  = "Lead Paint",
  "Superfund Proximity Summary Index_pct_usa"         = "Superfund",
  "RMP Proximity Summary Index_pct_usa"               = "RMP",
  "Hazardous Waste Proximity Summary Index_pct_usa"   = "Hazardous Waste",
  "Underground Storage Tanks Summary Index_pct_usa"   = "UST",
  "Wastewater Discharge Summary Index_pct_usa"        = "Wastewater",
  "Drinking Water Non-Compliance Summary Index_pct_usa" = "Drinking Water"
)
```

## 3.3 Harmonise and combine

```{r harmonise}
hbcu_scores <- ej_hbcu |>
  mutate(
    burden_score = rowMeans(across(all_of(pu_vars)), na.rm = TRUE),
    group        = "HBCU",
    year_founded = as.integer(year_founded)
  )

pwi_scores <- ej_pwi |>
  mutate(
    burden_score = rowMeans(across(all_of(pu_vars)), na.rm = TRUE),
    group        = "PWI",
    year_founded = as.integer(year_founded)
  )

combined <- bind_rows(hbcu_scores, pwi_scores) |>
  mutate(
    state_abbr = str_extract(Location, "[A-Z]{2}(?=[^A-Za-z]*$)"),
    region = case_when(
      state_abbr %in% c("CT","ME","MA","NH","NJ","NY","PA","RI","VT",
                        "MD","DE","DC")                              ~ "Northeast",
      state_abbr %in% c("AL","AR","FL","GA","KY","LA","MS","NC","OK",
                        "SC","TN","TX","VA","WV")                   ~ "South",
      state_abbr %in% c("IL","IN","IA","KS","MI","MN","MO","NE",
                        "ND","OH","SD","WI")                        ~ "Midwest",
      state_abbr %in% c("AK","AZ","CA","CO","HI","ID","MT","NV",
                        "NM","OR","UT","WA","WY")                   ~ "West",
      TRUE                                                           ~ "Other"
    )
  )

tibble(
  Group              = c("HBCU", "PWI"),
  N                  = c(nrow(hbcu_scores), nrow(pwi_scores)),
  `NA in burden_score` = c(
    sum(is.na(hbcu_scores$burden_score)),
    sum(is.na(pwi_scores$burden_score))
  )
) |>
  gt() |>
  tab_header(title = "Dataset summary after harmonisation")
```

------------------------------------------------------------------------

# 4 Descriptive Statistics

## 4.1 Summary statistics by group

```{r desc-table}
combined |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  group_by(group, indicator) |>
  summarise(
    N      = sum(!is.na(value)),
    Median = round(median(value, na.rm = TRUE), 1),
    Mean   = round(mean(value, na.rm = TRUE), 1),
    pct80  = round(mean(value >= 80, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  mutate(indicator = recode(indicator, !!!indicator_labels)) |>
  pivot_wider(
    names_from  = group,
    values_from = c(N, Median, Mean, pct80),
    names_glue  = "{group}_{.value}"
  ) |>
  arrange(indicator) |>
  gt() |>
  tab_header(
    title    = "EJ Summary Index USA percentiles: HBCU vs PWI",
    subtitle = "All institutions (pre-matching)"
  ) |>
  tab_spanner(label = "HBCU", columns = starts_with("HBCU")) |>
  tab_spanner(label = "PWI",  columns = starts_with("PWI"))  |>
  cols_label(
    indicator   = "Indicator",
    HBCU_N = "N", HBCU_Median = "Median", HBCU_Mean = "Mean", HBCU_pct80 = "% ≥ 80",
    PWI_N  = "N", PWI_Median  = "Median", PWI_Mean  = "Mean", PWI_pct80  = "% ≥ 80"
  ) |>
  tab_style(
    style     = cell_fill(color = "#fde0dd"),
    locations = cells_body(columns = HBCU_pct80, rows = HBCU_pct80 > 50)
  ) |>
  tab_style(
    style     = cell_fill(color = "#dbeafe"),
    locations = cells_body(columns = PWI_pct80, rows = PWI_pct80 > 50)
  ) |>
  tab_footnote(
    footnote  = "Red = > 50% of schools exceed the 80th national percentile.",
    locations = cells_column_labels(columns = HBCU_pct80)
  )
```

## 4.2 Indicator distributions

```{r violin-indicators, fig.height=10}
combined |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  mutate(
    indicator = recode(indicator, !!!indicator_labels),
    group     = factor(group, levels = c("HBCU", "PWI"))
  ) |>
  ggplot(aes(x = group, y = value, fill = group)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.15, outlier.size = 0.8) +
  geom_hline(yintercept = 50, linetype = "dashed",  colour = "grey50") +
  geom_hline(yintercept = 80, linetype = "dotted",  colour = "#E15759") +
  facet_wrap(~indicator, ncol = 4) +
  scale_fill_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), guide = "none"
  ) +
  labs(
    title    = "EJ Index USA percentile distributions: HBCU vs PWI",
    subtitle = "Dashed = 50th | dotted red = 80th national percentile",
    x = NULL, y = "USA percentile"
  )
```

------------------------------------------------------------------------

# 5 Testing Against the National Median

## 5.1 Normality (Shapiro–Wilk)

```{r normality}
combined |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  group_by(group, indicator) |>
  summarise(
    W      = round(shapiro.test(na.omit(value))$statistic, 4),
    p      = round(shapiro.test(na.omit(value))$p.value, 4),
    normal = if_else(p > 0.05, "✓ Normal", "✗ Non-normal"),
    .groups = "drop"
  ) |>
  mutate(indicator = recode(indicator, !!!indicator_labels)) |>
  pivot_wider(
    names_from  = group,
    values_from = c(W, p, normal),
    names_glue  = "{group}_{.value}"
  ) |>
  gt() |>
  tab_header(
    title    = "Shapiro–Wilk normality tests by group",
    subtitle = "H₀: data are normally distributed"
  ) |>
  tab_spanner(label = "HBCU", columns = starts_with("HBCU")) |>
  tab_spanner(label = "PWI",  columns = starts_with("PWI"))  |>
  cols_label(
    indicator    = "Indicator",
    HBCU_W = "W", HBCU_p = "p", HBCU_normal = "Normality?",
    PWI_W  = "W", PWI_p  = "p", PWI_normal  = "Normality?"
  ) |>
  tab_style(
    style     = cell_fill(color = "#fee8c8"),
    locations = cells_body(
      columns = c(HBCU_normal, PWI_normal),
      rows    = HBCU_normal == "✗ Non-normal" | PWI_normal == "✗ Non-normal"
    )
  )
```

## 5.2 One-sample Wilcoxon tests (H₁: median \> 50)

```{r one-sample-tests}
one_sample <- combined |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  group_by(group, indicator) |>
  summarise(
    n      = sum(!is.na(value)),
    median = round(median(value, na.rm = TRUE), 1),
    W      = round(wilcox.test(na.omit(value), mu = 50,
                               alternative = "greater")$statistic, 1),
    p      = round(wilcox.test(na.omit(value), mu = 50,
                               alternative = "greater")$p.value, 4),
    .groups = "drop"
  ) |>
  mutate(
    sig       = case_when(
      p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "ns"
    ),
    indicator = recode(indicator, !!!indicator_labels)
  ) |>
  pivot_wider(
    names_from  = group,
    values_from = c(n, median, W, p, sig),
    names_glue  = "{group}_{.value}"
  )

one_sample |>
  gt() |>
  tab_header(
    title    = "One-sample Wilcoxon tests vs. national median (μ = 50)",
    subtitle = "H₁: campus communities have above-median EJ burden nationally"
  ) |>
  tab_spanner(label = "HBCU", columns = starts_with("HBCU")) |>
  tab_spanner(label = "PWI",  columns = starts_with("PWI"))  |>
  cols_label(
    indicator   = "Indicator",
    HBCU_n = "N", HBCU_median = "Median", HBCU_W = "W",
    HBCU_p = "p", HBCU_sig = "",
    PWI_n  = "N", PWI_median  = "Median", PWI_W  = "W",
    PWI_p  = "p", PWI_sig  = ""
  ) |>
  tab_footnote(
    footnote  = "* p < 0.05  ** p < 0.01  *** p < 0.001 (one-sided, H₁: median > 50)",
    locations = cells_column_labels(columns = HBCU_sig)
  ) |>
  tab_style(
    style     = cell_fill(color = "#fde0dd"),
    locations = cells_body(columns = HBCU_p, rows = HBCU_p < 0.05)
  ) |>
  tab_style(
    style     = cell_fill(color = "#dbeafe"),
    locations = cells_body(columns = PWI_p, rows = PWI_p < 0.05)
  )
```

## 5.3 Forest plot: median percentiles with bootstrapped 95% CIs

```{r forest-plot, fig.height=7}
set.seed(RANDOM_SEED)
ci_both <- combined |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  group_by(group, indicator) |>
  summarise(
    median = median(value, na.rm = TRUE),
    lo = quantile(
      replicate(2000, median(sample(na.omit(value), replace = TRUE))), 0.025
    ),
    hi = quantile(
      replicate(2000, median(sample(na.omit(value), replace = TRUE))), 0.975
    ),
    .groups = "drop"
  ) |>
  mutate(
    indicator = recode(indicator, !!!indicator_labels),
    group     = factor(group, levels = c("HBCU", "PWI"))
  )

ggplot(ci_both, aes(x = median, y = reorder(indicator, median),
                    colour = group)) +
  geom_vline(xintercept = 50, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  geom_vline(xintercept = 80, linetype = "dotted",
             colour = "#E15759", linewidth = 0.7) +
  geom_errorbarh(
    aes(xmin = lo, xmax = hi), height = 0.3,
    position = position_dodge(0.6)
  ) +
  geom_point(size = 3, position = position_dodge(0.6)) +
  scale_colour_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), name = NULL
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title    = "Median EJ Index (USA percentile): HBCU vs PWI",
    subtitle = "Bootstrapped 95% CIs (2,000 replicates) | seed = 42",
    x = "USA percentile", y = NULL
  )
```

------------------------------------------------------------------------

# 6 EPA 80th-Percentile Threshold

```{r pct-80-chart, fig.height=6}
hbcu_pct <- hbcu_scores |>
  select(all_of(pu_vars)) |>
  pivot_longer(everything()) |>
  group_by(name) |>
  summarise(hbcu_pct80 = mean(value >= 80, na.rm = TRUE) * 100)

pwi_pct <- pwi_scores |>
  select(all_of(pu_vars)) |>
  pivot_longer(everything()) |>
  group_by(name) |>
  summarise(pwi_pct80 = mean(value >= 80, na.rm = TRUE) * 100)

pct_compare <- left_join(hbcu_pct, pwi_pct, by = "name") |>
  mutate(indicator = recode(name, !!!indicator_labels))

pct_compare |>
  pivot_longer(c(hbcu_pct80, pwi_pct80), names_to = "group",
               values_to = "pct") |>
  mutate(
    pct   = if_else(group == "pwi_pct80", -pct, pct),
    group = factor(group,
                   levels = c("hbcu_pct80", "pwi_pct80"),
                   labels = c("HBCU", "PWI"))
  ) |>
  ggplot(aes(x = pct, y = reorder(indicator, abs(pct)), fill = group)) +
  geom_col(alpha = 0.7) +
  scale_fill_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), name = NULL
  ) +
  scale_x_continuous(labels = abs) +
  labs(
    title    = "Percent of institutions ≥ 80th national percentile",
    subtitle = "HBCU (right) vs PWI (left)",
    x = "% institutions", y = NULL
  )
```

------------------------------------------------------------------------

# 7 Cumulative Burden Score

```{r burden-summary}
combined |>
  group_by(group) |>
  summarise(
    N       = n(),
    Min     = round(min(burden_score,               na.rm = TRUE), 1),
    Q1      = round(quantile(burden_score, .25,     na.rm = TRUE), 1),
    Median  = round(median(burden_score,            na.rm = TRUE), 1),
    Mean    = round(mean(burden_score,              na.rm = TRUE), 1),
    Q3      = round(quantile(burden_score, .75,     na.rm = TRUE), 1),
    Max     = round(max(burden_score,               na.rm = TRUE), 1),
    `% >= 70` = round(mean(burden_score >= 70,      na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(
    title    = "Cumulative EJ burden score by group",
    subtitle = "Mean of 13 EJ Summary Index USA percentiles"
  )
```

```{r burden-violin-all, fig.height=5}
ggplot(combined, aes(x = group, y = burden_score, fill = group)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.15, outlier.size = 1.5) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 3,
               fill = "white") +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey50") +
  scale_fill_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), guide = "none"
  ) +
  labs(
    title    = "Cumulative EJ burden score: HBCU vs PWI (all institutions)",
    subtitle = "Dashed = national median (50th percentile)",
    x = NULL, y = "Mean EJ Index USA percentile"
  )
```

------------------------------------------------------------------------

# 8 Propensity-Score Matching

1:1 nearest-neighbour matching on Census region and founding year. Ensures HBCU and PWI groups are comparable on key observable confounders before inferential comparison.

```{r matching}
match_df <- combined |>
  filter(!is.na(region), !is.na(year_founded)) |>
  mutate(treat = if_else(group == "HBCU", 1L, 0L))

set.seed(RANDOM_SEED)
m.out <- matchit(
  treat ~ region + year_founded,
  data   = match_df,
  method = "nearest",
  ratio  = 1
)

matched <- match.data(m.out)

summary(m.out)
```

```{r burden-violin-matched, fig.height=5}
wilcox_res <- wilcox.test(burden_score ~ group, data = matched)

matched |>
  mutate(group = factor(group, levels = c("HBCU", "PWI"))) |>
  ggplot(aes(x = group, y = burden_score, fill = group)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.15, outlier.size = 1.5) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 3,
               fill = "white") +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey50") +
  scale_fill_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), guide = "none"
  ) +
  labs(
    title    = "Cumulative EJ burden: HBCU vs matched PWI",
    subtitle = paste0(
      "Two-sample Wilcoxon: W = ", wilcox_res$statistic,
      ",  p = ", formatC(wilcox_res$p.value, format = "e", digits = 2)
    ),
    x = NULL, y = "Mean EJ Index USA percentile"
  )
```

------------------------------------------------------------------------

# 9 Indicator-Level Comparison (Matched Sample)

```{r indicator-tests}
indicator_tests <- matched |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  group_by(indicator) |>
  summarise(
    HBCU_median = round(median(value[group == "HBCU"], na.rm = TRUE), 1),
    PWI_median  = round(median(value[group == "PWI"],  na.rm = TRUE), 1),
    difference  = round(HBCU_median - PWI_median, 1),
    W           = round(wilcox.test(value[group == "HBCU"],
                                    value[group == "PWI"])$statistic, 1),
    p           = round(wilcox.test(value[group == "HBCU"],
                                    value[group == "PWI"])$p.value, 4),
    .groups     = "drop"
  ) |>
  mutate(
    sig       = case_when(
      p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "ns"
    ),
    indicator = recode(indicator, !!!indicator_labels)
  ) |>
  arrange(p)

indicator_tests |>
  gt() |>
  tab_header(
    title    = "Two-sample Wilcoxon tests per indicator (matched sample)",
    subtitle = "H₁: HBCU campus communities differ from matched PWI communities"
  ) |>
  cols_label(
    indicator   = "Indicator",
    HBCU_median = "HBCU Median",
    PWI_median  = "PWI Median",
    difference  = "Difference",
    W = "W", p = "p-value", sig = ""
  ) |>
  tab_footnote(
    footnote  = "Positive difference = HBCU median exceeds PWI median. * p<0.05  ** p<0.01  *** p<0.001",
    locations = cells_column_labels(columns = sig)
  ) |>
  tab_style(
    style     = cell_fill(color = "#fde0dd"),
    locations = cells_body(rows = p < 0.05)
  ) |>
  tab_style(
    style     = cell_fill(color = "#fee8c8"),
    locations = cells_body(rows = p >= 0.05 & p < 0.10)
  )
```

```{r indicator-forest, fig.height=7}
set.seed(RANDOM_SEED)
indicator_ci_df <- matched |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  group_by(group, indicator) |>
  summarise(
    median = median(value, na.rm = TRUE),
    lo = quantile(
      replicate(2000, median(sample(na.omit(value), replace = TRUE))), 0.025
    ),
    hi = quantile(
      replicate(2000, median(sample(na.omit(value), replace = TRUE))), 0.975
    ),
    .groups = "drop"
  ) |>
  mutate(
    indicator = recode(indicator, !!!indicator_labels),
    group     = factor(group, levels = c("HBCU", "PWI"))
  )

ggplot(indicator_ci_df,
       aes(x = median, y = reorder(indicator, median), colour = group)) +
  geom_vline(xintercept = 50, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  geom_vline(xintercept = 80, linetype = "dotted",
             colour = "#E15759", linewidth = 0.7) +
  geom_errorbarh(
    aes(xmin = lo, xmax = hi), height = 0.3,
    position = position_dodge(0.6)
  ) +
  geom_point(size = 3, position = position_dodge(0.6)) +
  scale_colour_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), name = NULL
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title    = "Median EJ Index (USA percentile): matched HBCU vs PWI",
    subtitle = "Bootstrapped 95% CIs | dashed = 50th | dotted red = 80th percentile",
    x = "USA percentile", y = NULL
  )
```

------------------------------------------------------------------------

# 10 Geographic Distribution

```{r map-institutions, fig.height=7}
library(sf)
library(tigris)

hbcu_coords <- readr::read_csv("hbcu_list.csv", show_col_types = FALSE) |>
  select(school, Latitude, Longitude)
pwi_coords  <- readr::read_csv("pwi_list.csv",  show_col_types = FALSE) |>
  select(school, Latitude, Longitude)

all_institutions <- bind_rows(
  hbcu_scores |> select(school, burden_score, group) |>
    left_join(hbcu_coords, by = "school"),
  pwi_scores  |> select(school, burden_score, group) |>
    left_join(pwi_coords,  by = "school")
) |>
  filter(!is.na(Latitude), !is.na(Longitude)) |>
  mutate(group = factor(group, levels = c("HBCU", "PWI")))

states_sf <- tigris::states(cb = TRUE, resolution = "20m",
                            progress_bar = FALSE) |>
  filter(!STUSPS %in% c("AK","HI","PR","GU","VI","MP","AS")) |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = states_sf, fill = "grey96", colour = "grey70",
          linewidth = 0.3) +
  geom_point(
    data  = filter(all_institutions, !is.na(Longitude)),
    aes(x = Longitude, y = Latitude, colour = group, size = burden_score),
    alpha = 0.75
  ) +
  scale_colour_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"),
    name   = "Institution type"
  ) +
  scale_size_continuous(range = c(1.5, 5), name = "EJ burden\nscore") +
  coord_sf(xlim = c(-107, -67), ylim = c(24, 48)) +
  labs(
    title    = "Geographic distribution of HBCU and PWI institutions",
    subtitle = "Point size proportional to cumulative EJ burden score",
    x = NULL, y = NULL
  ) +
  guides(
    colour = guide_legend(override.aes = list(size = 4)),
    size   = guide_legend(override.aes = list(colour = "grey40"))
  ) +
  theme(
    axis.text        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid.major = element_line(colour = "grey90")
  )
```

```{r leaflet-map}
pal <- colorFactor(
  palette = c("#B2182B", "#4E79A7"),
  domain  = all_institutions$group
)

leaflet(all_institutions) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng         = ~Longitude,
    lat         = ~Latitude,
    radius      = ~ scales::rescale(burden_score, to = c(4, 12)),
    color       = ~ pal(group),
    fillColor   = ~ pal(group),
    fillOpacity = 0.8,
    stroke      = TRUE,
    weight      = 1,
    opacity     = 0.9,
    popup       = ~ paste0(
      "<b>", school, "</b><br>",
      "Type: <b>", group, "</b><br>",
      "EJ burden score: <b>", round(burden_score, 1), "</b>"
    )
  ) |>
  addLegend(
    pal = pal, values = ~group,
    title = "Institution type", position = "bottomright"
  )
```

------------------------------------------------------------------------

# 11 Within-Group Variability

## 11.1 Variance and IQR

```{r variance-table}
combined |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  group_by(group, indicator) |>
  summarise(
    SD  = round(sd(value,   na.rm = TRUE), 1),
    IQR = round(IQR(value,  na.rm = TRUE), 1),
    CV  = round(sd(value,   na.rm = TRUE) /
                mean(value, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  mutate(indicator = recode(indicator, !!!indicator_labels)) |>
  pivot_wider(
    names_from  = group,
    values_from = c(SD, IQR, CV),
    names_glue  = "{group}_{.value}"
  ) |>
  arrange(desc(HBCU_IQR)) |>
  gt() |>
  tab_header(
    title    = "Within-group variability in EJ Index USA percentiles",
    subtitle = "SD = standard deviation | IQR = interquartile range | CV = coefficient of variation (%)"
  ) |>
  tab_spanner(label = "HBCU", columns = starts_with("HBCU")) |>
  tab_spanner(label = "PWI",  columns = starts_with("PWI"))  |>
  cols_label(
    indicator = "Indicator",
    HBCU_SD = "SD", HBCU_IQR = "IQR", HBCU_CV = "CV",
    PWI_SD  = "SD", PWI_IQR  = "IQR", PWI_CV  = "CV"
  )
```

## 11.2 Levene's test

```{r levene-tests}
library(car)

levene_res <- map_dfr(pu_vars, function(v) {
  df <- combined |>
    select(group, value = all_of(v)) |>
    filter(!is.na(value))
  lt <- leveneTest(value ~ factor(group), data = df, center = median)
  tibble(
    indicator = v,
    F_stat    = round(lt$`F value`[1], 3),
    p         = round(lt$`Pr(>F)`[1], 4)
  )
}) |>
  mutate(
    sig = case_when(
      p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "ns"
    ),
    indicator = recode(indicator, !!!indicator_labels),
    more_variable = if_else(
      p < 0.05,
      combined |>
        select(group, all_of(pu_vars)) |>
        pivot_longer(-group, names_to = "ind") |>
        group_by(group, ind) |>
        summarise(sd = sd(value, na.rm = TRUE), .groups = "drop") |>
        pivot_wider(names_from = group, values_from = sd) |>
        mutate(more = if_else(HBCU > PWI, "HBCU", "PWI")) |>
        pull(more),
      "-"
    )
  )

levene_res |>
  gt() |>
  tab_header(
    title    = "Levene’s test: equality of variance between groups",
    subtitle = "H₀: variance is equal | centre = median"
  ) |>
  cols_label(
    indicator     = "Indicator",
    F_stat        = "F",
    p             = "p-value",
    sig           = "",
    more_variable = "More variable group"
  ) |>
  tab_style(
    style     = cell_fill(color = "#fde0dd"),
    locations = cells_body(rows = p < 0.05 & more_variable == "HBCU")
  ) |>
  tab_style(
    style     = cell_fill(color = "#dbeafe"),
    locations = cells_body(rows = p < 0.05 & more_variable == "PWI")
  )
```

## 11.3 Ridgeline distribution plot

```{r ridgeline, fig.height=5}
library(ggridges)

combined |>
  select(group, all_of(pu_vars)) |>
  pivot_longer(-group, names_to = "indicator") |>
  mutate(
    indicator = recode(indicator, !!!indicator_labels),
    group     = factor(group, levels = c("PWI", "HBCU"))
  ) |>
  ggplot(aes(x = value, y = indicator, fill = group)) +
  geom_density_ridges(
    alpha = 0.55, scale = 0.9, rel_min_height = 0.01,
    colour = "white", linewidth = 0.3
  ) +
  geom_vline(xintercept = 50, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 80, linetype = "dotted", colour = "#E15759") +
  scale_fill_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), name = NULL
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title    = "EJ burden distribution by indicator: HBCU vs PWI",
    subtitle = "Dashed = national median (50) | dotted red = EPA threshold (80)",
    x = "USA percentile", y = NULL
  ) +
  theme(legend.position = "top")
```

------------------------------------------------------------------------

# 12 Regional Comparison

```{r regional-comparison, fig.height=6}
region_sum <- combined |>
  filter(!is.na(region)) |>
  group_by(region, group) |>
  summarise(
    n      = n(),
    Median = round(median(burden_score, na.rm = TRUE), 1),
    Mean   = round(mean(burden_score,   na.rm = TRUE), 1),
    SD     = round(sd(burden_score,     na.rm = TRUE), 1),
    .groups = "drop"
  )

region_sum |>
  gt() |>
  tab_header(
    title    = "Cumulative EJ burden score by region and institution type",
    subtitle = "Mean of 13 EJ Summary Index USA percentiles"
  ) |>
  cols_label(
    region = "Region", group = "Type", n = "N",
    Median = "Median", Mean = "Mean", SD = "SD"
  )
```

```{r regional-violin, fig.height=6}
combined |>
  filter(!is.na(region)) |>
  mutate(
    group  = factor(group,  levels = c("HBCU", "PWI")),
    region = factor(region, levels = c("South", "Northeast",
                                       "Midwest", "West", "Other"))
  ) |>
  ggplot(aes(x = group, y = burden_score, fill = group)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.size = 1) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey50") +
  facet_wrap(~region, nrow = 1) +
  scale_fill_manual(
    values = c("HBCU" = "#B2182B", "PWI" = "#4E79A7"), guide = "none"
  ) +
  labs(
    title    = "Cumulative EJ burden score by Census region",
    subtitle = "HBCU vs PWI within each region | dashed = national median",
    x = NULL, y = "EJ burden score"
  )
```

```{r regional-wilcox}
regional_tests <- combined |>
  filter(!is.na(region)) |>
  group_by(region) |>
  filter(n_distinct(group) == 2) |>
  summarise(
    n_hbcu   = sum(group == "HBCU"),
    n_pwi    = sum(group == "PWI"),
    med_hbcu = round(median(burden_score[group == "HBCU"], na.rm = TRUE), 1),
    med_pwi  = round(median(burden_score[group == "PWI"],  na.rm = TRUE), 1),
    diff     = round(med_hbcu - med_pwi, 1),
    W        = wilcox.test(burden_score[group == "HBCU"],
                           burden_score[group == "PWI"])$statistic,
    p        = round(wilcox.test(burden_score[group == "HBCU"],
                                 burden_score[group == "PWI"])$p.value, 4),
    .groups  = "drop"
  ) |>
  mutate(sig = case_when(
    p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "ns"
  ))

regional_tests |>
  gt() |>
  tab_header(
    title    = "Within-region Wilcoxon tests: HBCU vs PWI burden score",
    subtitle = "Positive diff = HBCU burden > PWI burden"
  ) |>
  cols_label(
    region   = "Region",
    n_hbcu   = "N (HBCU)", n_pwi  = "N (PWI)",
    med_hbcu = "HBCU Median", med_pwi = "PWI Median",
    diff     = "Difference", W = "W", p = "p-value", sig = ""
  ) |>
  tab_style(
    style     = cell_fill(color = "#fde0dd"),
    locations = cells_body(rows = p < 0.05)
  )
```

------------------------------------------------------------------------

# 13 Effect Sizes

Three complementary effect size measures: **Cohen's d** (standardised mean difference); **rank-biserial correlation r** (non-parametric, derived from Wilcoxon W); **Common Language Effect Size (CLES)** — the probability that a randomly selected HBCU community exceeds a randomly selected PWI community.

```{r effect-sizes}
cohens_d <- function(x, y) {
  n1 <- sum(!is.na(x)); n2 <- sum(!is.na(y))
  sp <- sqrt(((n1-1)*var(x, na.rm=TRUE) + (n2-1)*var(y, na.rm=TRUE)) /
               (n1 + n2 - 2))
  (mean(x, na.rm=TRUE) - mean(y, na.rm=TRUE)) / sp
}
rb_corr <- function(W, n1, n2) 1 - (2 * W) / (n1 * n2)
cles    <- function(x, y) mean(outer(x, y, ">"), na.rm = TRUE)

effect_df <- map_dfr(pu_vars, function(v) {
  hv <- hbcu_scores[[v]]; pv <- pwi_scores[[v]]
  wt <- wilcox.test(hv, pv, na.action = na.omit)
  n1 <- sum(!is.na(hv)); n2 <- sum(!is.na(pv))
  tibble(
    indicator = v,
    d    = round(cohens_d(hv, pv), 3),
    r    = round(rb_corr(wt$statistic, n1, n2), 3),
    CLES = round(cles(na.omit(hv), na.omit(pv)) * 100, 1)
  )
}) |>
  mutate(
    indicator = recode(indicator, !!!indicator_labels),
    d_label = case_when(
      abs(d) >= 0.8 ~ "Large", abs(d) >= 0.5 ~ "Medium",
      abs(d) >= 0.2 ~ "Small", TRUE ~ "Negligible"
    ),
    r_label = case_when(
      abs(r) >= 0.5 ~ "Large", abs(r) >= 0.3 ~ "Medium",
      abs(r) >= 0.1 ~ "Small", TRUE ~ "Negligible"
    )
  ) |>
  arrange(desc(abs(d)))

effect_df |>
  gt() |>
  tab_header(
    title    = "Effect sizes: HBCU vs PWI EJ burden per indicator",
    subtitle = "Positive d and r indicate HBCU communities are more burdened"
  ) |>
  cols_label(
    indicator = "Indicator",
    d = "Cohen’s d", d_label = "d magnitude",
    r = "Rank-biserial r", r_label = "r magnitude",
    CLES = "CLES (%)"
  ) |>
  tab_footnote(
    footnote  = "CLES = P(HBCU community > PWI community on this indicator).",
    locations = cells_column_labels(columns = CLES)
  ) |>
  tab_style(
    style     = cell_fill(color = "#fde0dd"),
    locations = cells_body(
      columns = d_label,
      rows    = d_label %in% c("Large", "Medium")
    )
  ) |>
  tab_style(
    style     = cell_fill(color = "#fde0dd"),
    locations = cells_body(columns = CLES, rows = CLES > 65)
  )
```

```{r effect-dotplot, fig.height=6}
effect_df |>
  ggplot(aes(x = r, y = reorder(indicator, r))) +
  geom_vline(xintercept = 0,   colour = "grey50",  linewidth = 0.6) +
  geom_vline(xintercept = 0.3, linetype = "dashed",
             colour = "grey60", linewidth = 0.4) +
  geom_vline(xintercept = 0.5, linetype = "dotted",
             colour = "#E15759", linewidth = 0.5) +
  geom_segment(aes(xend = 0, yend = indicator), colour = "grey80") +
  geom_point(aes(colour = r), size = 4) +
  scale_colour_gradient2(
    low      = "#4E79A7", mid = "grey80", high = "#B2182B",
    midpoint = 0, name = "Rank-biserial r"
  ) +
  scale_x_continuous(limits = c(-0.1, 1), breaks = seq(0, 1, 0.25)) +
  labs(
    title    = "Rank-biserial effect size: HBCU vs PWI per indicator",
    subtitle = "Dashed = medium (0.3) | dotted = large (0.5) | positive = HBCU more burdened",
    x = "Rank-biserial correlation (r)", y = NULL
  )
```

------------------------------------------------------------------------

# 14 Save Outputs

```{r save-outputs}
# Save the matched dataset for the Dryad repository
write_csv(matched, "matched_hbcu_pwi.csv")

# Session information for full reproducibility record
sessionInfo()
```